---
title: "Lab 1 - Data visualization"
format: pdf
editor: visual
author: "Luke Owen"
---

```{r load tidyverse}
library(tidyverse)
```

# Questions

## Part 1

### Question 1

Binwidth 100

```{r #histogram of midwest, bin = 100}

ggplot(midwest,aes(x = popdensity)) + 
         geom_histogram(binwidth = 100) + 
         labs( 
           x = "Population Density",
           y = "Count",
           title = "Population density of midwestern counties", 
           subtitle = "Binwidth = 100")

```
comment on binwidth: Too small based on data range. the histogram becomes noisy and overemphasizes minor fluctuations rather than the overall distribution.


Binwidth 1,000

```{r}
#histogram of midwest, bin = 1000
ggplot(midwest,aes(x = popdensity)) + 
         geom_histogram(binwidth = 1000) + 
         labs( 
           x = "Population Density",
           y = "Count",
           title = "Population density of midwestern counties", 
           subtitle = "Binwidth = 1000")

```
comment on binwidth: Best balance. Shows the right-skewed distribution and concentration of counties at low population densities while retaining meaningful structure viewing.


Binwidth: 10,000

```{r}
#histogram of midwest, bin = 10000
ggplot(midwest,aes(x = popdensity)) + 
         geom_histogram(binwidth = 10000) + 
         labs( 
           x = "Population Density",
           y = "Count",
           title = "Population density of midwestern counties", 
           subtitle = "Binwidth = 10000")


```
comment on binwidth: Too coarse. Important variation among lower-density counties is lost.


Binwidth 100,000

```{r}
#histogram of midwest, bin = 100,000
ggplot(midwest,aes(x = popdensity)) + 
         geom_histogram(binwidth = 100000) + 
         labs( 
           x = "Population Density",
           y = "Count",
           title = "Population density of midwestern counties", 
           subtitle = "Binwidth = 100000")


```
comment on binwidth: Far too coarse. Most counties collapse into very few bins, obscuring the distribution almost entirely.

### Question 2

### Question 3

### Question 4

### Question 5

### Question 6

### Question 7

## Part 2

**Enough about the Midwest!**

### Question 8

### Question 9
